\documentclass[11pt]{article}
% Compile with: pdflatex outline.tex (run twice for cross-references/ToC)
%
% NOTE ON DOCUMENT CLASS: Automated Software Engineering (Springer) submissions ultimately require
% Springer Nature's official class (e.g. sn-jnl.cls, "sn-basic" / numbered-reference layout), which
% is not vendored in this repository. This outline uses the standard article class as a portable
% stand-in so it compiles anywhere; swap in Springer's class and \maketitle-equivalent frontmatter
% commands when the official template is downloaded from the journal's submission site, mirroring
% how docs/research/jss/outline.tex uses elsarticle for the Elsevier-targeted companion submission.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage[margin=1in]{geometry}

\title{Graph-Based Detection of Architectural Anti-Patterns and Prescriptive Refactoring in
Distributed Publish--Subscribe Systems}
\author{Author Name(s) Here}
\date{}

\begin{document}
\maketitle

\begin{center}
\emph{Target: Automated Software Engineering (AuSE), Springer --- Special Issue on
``Intelligent techniques for CI/CD, DevOps, software evolution, technical debt analysis, and
refactoring recommendation.''}

\emph{SI topic mapping: this submission covers three of the bullet's five named areas directly ---
\textbf{technical debt analysis} (the anti-pattern catalog, \S4), \textbf{refactoring recommendation}
(the prescriptive operators, \S6), and \textbf{CI/CD/DevOps} (the delta-aware quality gate, \S7). No
AutoML/NAS/LLM contribution is claimed.}
\end{center}

\bigskip

\noindent\textbf{Provenance note (not part of the manuscript).} This outline revises the prior
SaG-Prescribe outline to make \textbf{anti-pattern detection} a first-class, explicitly evaluated
contribution rather than an input inherited from a companion diagnostic paper. The catalog of
twenty-one publish--subscribe anti-patterns and its empirical validation
(\texttt{docs/antipatterns.md}) already exist as a mature, separately maintained artifact in the
source repository but were not previously folded into an AuSE-targeted manuscript; this revision does
so, pairing them with the existing closed-loop prescriptive-refactoring pipeline
(\texttt{docs/prescription.md}, \texttt{docs/research/ause/draft.md}) under one narrative:
\emph{detect named, severity-tiered architectural anti-patterns $\rightarrow$ prescribe targeted,
verified refactorings $\rightarrow$ gate CI/CD on newly introduced regressions.} The full manuscript
draft (\texttt{draft.md}) still reflects the prior prescription-only framing and has not yet been
rewritten to match this outline --- see the pre-submission checklist at the end of this document.

\bigskip
\hrule
\bigskip

\begin{abstract}
Distributed publish--subscribe middleware (ROS~2, DDS, MQTT) decouples producers and consumers, but
the resulting indirect dependency structure obscures how component failures cascade, and this
architectural technical debt accumulates invisibly to code-level static analysis (SCA) tools. Unlike
object-oriented design, where mature catalogs of named anti-patterns (God Class, Feature Envy,
Shotgun Surgery) give practitioners a shared vocabulary and testable detection rules, distributed
publish--subscribe architectures have no equivalent catalog: problems are discovered reactively,
through postmortems and cascade incidents, rather than proactively at design time. Even where
structural diagnostic frameworks exist, they typically operate \emph{open-loop}: they rank components
by criticality without naming the specific architectural pathology at fault or producing verified
guidance on how to repair it. We address both gaps with \textbf{SaG-Prescribe}, a graph-based
framework that (1) detects twenty-one named, severity-tiered publish--subscribe anti-patterns and bad
smells --- including Single Point of Failure, God Component, Broker Saturation (Hub-and-Spoke), Chatty
Pair, and QoS Policy Mismatch --- via formally specified topological signatures over thirteen
structural metrics and adaptive box-plot thresholds, each validated against independent
failure-simulation ground truth (Spearman $\rho = 0.876$ overall, rising to $\rho = 0.943$ at large
scale; $F_1 = 0.923$, precision $0.912$, recall $0.857$ for critical-tier classification); and (2)
compiles the resulting diagnosis --- components and patterns flagged \texttt{CRITICAL} or
\texttt{HIGH} --- into a transformation policy $\Delta(G)$ of three targeted graph-mutation operators
(logical topic splitting, physical anti-affinity reallocation, and transport QoS contract hardening),
each re-verified by the same discrete-event cascade simulator that produced the diagnosis before it
is surfaced as a recommendation. The pipeline is further operationalized as a delta-aware CI/CD
quality gate (\texttt{detect\_antipatterns.py}) that blocks only newly introduced structural
anti-patterns relative to a Git merge base, within sub-minute execution bounds.

Across eight benchmark scenarios used for detection validation and a seven-scenario subset used for
prescriptive evaluation --- spanning autonomous vehicles, IoT, finance, healthcare, and hyper-scale
enterprise topologies --- the generated prescriptions reduce the System Risk Index (SRI, lower is
better) in every scenario, with relative risk reductions between 1.4\% and 15.9\% (Wilcoxon
signed-rank across the seven baseline/mutated pairs: $W=0.0$, $p=0.0156$, significant at
$\alpha=0.05$). We report this result alongside a less flattering but more honest one: component-level
failure-impact reductions average only $+4.61\%$ and regress in two of seven scenarios ($-31.67\%$ and
$-25.36\%$), because the current prescriptive policy is applied greedily and unconditionally,
occasionally introducing new cascade hops that offset its own gains --- a finding that directly
motivates a per-edit acceptance filter as the highest-priority extension. The full generate--verify
loop completes in under 65 seconds for six of the seven prescription scenarios (4.7\,s--64.3\,s) and
in approximately 10.8 minutes for the 300-process enterprise topology, while the standalone CI/CD
detection gate completes in under 2 seconds to $\sim$40 seconds across the same scale range --- making
both proactive anti-pattern detection and closed-loop prescriptive refactoring practical as
merge-request-gated or nightly pre-deployment CI/CD stages.
\end{abstract}

\noindent\textbf{Keywords:} architectural anti-patterns; bad smells; technical debt analysis;
refactoring recommendation; publish--subscribe middleware; CI/CD quality gates; DevOps; failure
cascade simulation; graph mutation; search-based software engineering

\bigskip
\hrule
\bigskip

% ---------------------------------------------------------------------------
% MANUSCRIPT OUTLINE
% This file captures the outline (title, abstract, keywords, and annotated
% section/subsection structure) for editorial review before full drafting.
% Each \paragraph{} below is a planned subsection with a one- or two-sentence
% description of its content, not the final prose.
% ---------------------------------------------------------------------------

\section{Introduction}

\paragraph{1.1 Context and motivation}
Pub-sub decoupling as the backbone of cyber-physical/microservice/IoT systems; decoupling obscures
failure propagation and the resulting architectural technical debt.

\paragraph{1.2 Two open gaps}
\textbf{(a)} The Architecture-Code Gap and the absence of a named, testable anti-pattern catalog for
pub-sub topologies (SCA tools are topology-blind; OO/microservices catalogs don't transfer to
broker-mediated communication); \textbf{(b)} the open-loop refactoring-recommendation gap --- even
topology-aware diagnostics (including our own prior SaG framework) rank criticality without naming the
fault or verifying its remediation.

\paragraph{1.3 Proposed solution}
A unified detect$\to$prescribe$\to$gate pipeline --- twenty-one-pattern catalog with empirically
validated detection rules, feeding a closed-loop mutation-and-resimulation engine, operationalized as
a delta-aware CI/CD gate.

\paragraph{1.4 Contributions}
Four explicit, numbered items: (1) the empirically-validated anti-pattern catalog and detection
methodology; (2) the closed-loop prescriptive pipeline and its three operators, explicitly mapped to
the patterns they repair; (3) the CI/CD gate; (4) the multi-scenario evaluation, including the honest
component-level result.

\section{Background and Related Work}

\paragraph{2.1 Publish--subscribe middleware dependability}
Broker fault tolerance, Kafka, DDS QoS/latency literature.

\paragraph{2.2 Anti-pattern and code-smell catalogs}
Object-oriented design (Fowler's refactoring catalog; Brown et al.'s architectural AntiPatterns;
Suryanarayana et al.'s design smells) and microservices smells (Richardson; Taibi et al.), establishing
the template this catalog follows --- named pattern, formal detection rule, refactoring strategy ---
while operating on system \emph{topology} rather than code or REST call graphs.

\paragraph{2.3 Refactoring recommendation and architectural technical debt}
Code-scope recommenders (smell-driven, learning-based, LLM-based) vs. this paper's topology scope and
verify-before-recommend model. \emph{(Two of three prior citation slots now fillable from \S2.2's
real references; learning-based and LLM-based recommender slots remain pending --- no invented
references.)}

\paragraph{2.4 Search-Based Software Engineering and architecture optimization}
Open-loop vs. this paper's closed-loop verification.

\paragraph{2.5 Diagnostic foundation (SaG)}
The heterogeneous graph model and RMAV attribution this paper builds on; what is summarized vs. not
repeated from the companion diagnostic paper.

\paragraph{2.6 Structural criticality analysis}
Why classical single-metric centrality (degree, betweenness alone) degrades on typed pub-sub graphs
relative to the composite $Q(v)$ score.

\section{System Model and Code-Quality-Augmented Quality Attribution}

\paragraph{3.1 Heterogeneous graph formulation}
Five node types, six structural edge types.

\paragraph{3.2 Derived \texttt{DEPENDS\_ON} projection}
App$\leftrightarrow$App, App$\leftrightarrow$Broker, App$\leftrightarrow$Library,
Broker$\leftrightarrow$Broker, and the four architectural layers (app, infra, mw, system).

\paragraph{3.3 Code Quality Penalty (CQP)}
The paper's explicit bridge from SCA metrics to architecture-level risk; feeds the Maintainability
dimension below.

\paragraph{3.4 Multi-dimensional quality attribution (RMAV)}
Reliability/Maintainability/Availability/Vulnerability, AHP-weighted composite $Q(v)$, adaptive
box-plot criticality tiers. This section is the shared foundation consumed by both the detection
catalog (\S4) and the prescriptive engine (\S6).

\section{A Catalog of Architectural Anti-Patterns for Publish--Subscribe Systems}
\emph{(new detection contribution)}

\paragraph{4.1 Anti-patterns vs. bad smells}
Taxonomy and confidence distinction (structural configurations with well-understood failure modes vs.
heuristic surface symptoms requiring human judgment).

\paragraph{4.2 Detection methodology}
Thirteen structural metrics per component (PageRank, Reverse PageRank, betweenness, closeness,
eigenvector centrality, in/out-degree, clustering coefficient, articulation point score, bridge ratio,
QoS-weighted degree measures); rank-based vs. min-max normalization; adaptive box-plot thresholds
(scale-invariant, distribution-aware, statistically grounded) in place of fixed global constants.

\paragraph{4.3 Catalog overview}
Twenty-one patterns organized into three severity tiers (\texttt{CRITICAL}, \texttt{HIGH},
\texttt{MEDIUM}) across four RMAV dimensions; summary table of pattern, severity, primary RMAV
dimension, and formal detection rule (full specifications and remediation strategies given in the
companion technical reference, \texttt{docs/antipatterns.md}, cited rather than reproduced in full).

\paragraph{4.4 Representative pattern walkthroughs}
Five patterns selected to span severity tiers, RMAV dimensions, and detection technique diversity:
\texttt{SPOF} (articulation-point-based, Availability), \texttt{GOD\_COMPONENT} (betweenness +
maintainability gate), \texttt{BROKER\_OVERLOAD}/Hub-and-Spoke (the pub-sub instantiation of a
classical topology anti-pattern, validated on a deliberately-encoded scenario), \texttt{CHATTY\_PAIR}
(bidirectional edge-weight product, hidden coupling behind nominal decoupling), and
\texttt{QOS\_MISMATCH} (transport-contract-boundary detection unique to QoS-bearing middleware).

\paragraph{4.5 Empirical validation of detection}
Validation methodology (simulated failure impact $I(v)$ as ground truth); headline metrics (Spearman
$\rho(Q,I) = 0.876$ overall, $0.943$ at large scale; $F_1 = 0.923$, precision $0.912$, recall $0.857$,
Top-5 overlap $0.80$); pattern-specific evidence (SPOF $F_1 > 0.95$; Hub-and-Spoke scenario
confirmation); baseline comparison against single-metric centrality; the eight-scenario validation
suite, foregrounding Scenario~06 (sparse microservices mesh) as the \textbf{precision stress test} ---
a well-structured topology should produce few or no findings --- and Scenario~07 as the
\textbf{scalability benchmark}.

\section{Closed-Loop Optimization Objective}

Formal $\min_\Delta$ objective under a (currently unconstrained) modification budget:
\begin{equation*}
\min_{\Delta} \sum_{v \in V} I^*_{\Delta(G)}(v) \quad \text{subject to} \quad \mathrm{Cost}(\Delta) \le \mathcal{B}
\end{equation*}
Acceptance criterion $\Delta\mathrm{SRI}>0$ as implemented in \texttt{PrescribeService}; explicit
statement that the candidate set for $\Delta$ is exactly the \texttt{CRITICAL}/\texttt{HIGH}
components and patterns identified in \S4; note on the deferred margin-based criterion.

\section{The SaG-Prescribe Prescriptive Pipeline}

\paragraph{6.1 Hexagonal core abstraction}
\texttt{IGraphRepository}, \texttt{Neo4jRepository} vs. \texttt{MemoryRepository}.

\paragraph{6.2 Pipeline stages 1--7}
Diagnostic foundation and anti-pattern detection $\to$ prescriptive generation $\to$ review interface;
explicit hand-off from \S4's named findings to operator selection.

\paragraph{6.3 Three refactoring operators}
Logical topic splitting, physical anti-affinity reallocation, transport QoS contract hardening ---
each formalized as a typed graph mutation rule and explicitly mapped to the anti-patterns it targets:
topic splitting $\to$ \texttt{TOPIC\_FANOUT}/\texttt{GOD\_COMPONENT}; anti-affinity reallocation $\to$
\texttt{SPOF}/\texttt{BROKER\_OVERLOAD}/\texttt{HUB\_AND\_SPOKE}; QoS hardening $\to$
\texttt{QOS\_MISMATCH} and QoS-fragile \texttt{CRITICAL}/\texttt{HIGH} channels.

\paragraph{6.4 Closed-loop verification procedure}
Export $\to$ mutate $\to$ sandbox $\to$ resimulate $\to$ $\Delta$SRI, with the independence guarantee
against circular leakage.

\section{DevOps Integration and Delta-Aware CI/CD Gating}
\emph{(This section is the paper's most direct link to the SI's CI/CD/DevOps bullet --- kept
prominent.)}

\paragraph{7.1 Automated code review architecture}
\texttt{detect\_antipatterns.py} surfacing \S4's twenty-one detectors as a blocking CI check.

\paragraph{7.2 Delta-aware regression semantics}
Merge-base diffing, waiver register for accepted legacy risk.

\paragraph{7.3 Exit-code protocol}
0 (pass) / 1 (warn) / 2 (block).

\section{Experimental Design}

\paragraph{8.1 Research questions}
\begin{description}[leftmargin=2.2cm,style=nextline]
  \item[RQ1 (Detection efficacy and precision)] Does the catalog correlate with ground-truth impact,
    and does it avoid over-flagging well-structured systems?
  \item[RQ2 (Prescriptive efficacy)] Does the closed-loop engine reduce the System Risk Index across
    heterogeneous scenarios, and are the reductions statistically significant?
  \item[RQ3 (Operator contributions)] How do the individual refactoring operators contribute to the
    observed improvements across topological regimes?
  \item[RQ4 (Component-level remediation efficacy)] Do system-level SRI improvements translate
    uniformly into component-level failure-impact reductions, or can the greedy, unconditional policy
    application backfire on individual components?
  \item[RQ5 (Computational overhead and CI/CD feasibility)] What is the wall-clock execution time of
    the detection gate and of the full prescriptive pipeline as system scale grows, and is either
    compatible with CI/CD budgets?
\end{description}

\paragraph{8.2 Scenario suites}
Eight scenarios (01--08) for detection validation, including the deterministic ``Tiny Regression''
smoke-test fixture (\S4.5); the seven-scenario subset (01--07, autonomous vehicle $\to$ hyper-scale
enterprise) used for prescriptive evaluation, explicitly noting and justifying the count mismatch (the
smoke-test fixture carries no domain-representative signal for prescriptive evaluation).

\paragraph{8.3 Experimental protocol}
Deterministic simulator, verified $\sigma_{\text{seed}}=0$ across 5 seeds.

\paragraph{8.4 Metrics}
Detection: Spearman $\rho$, $F_1$/precision/recall, Top-$k$ overlap. Prescription: SRI formula,
$\Delta$SRI, operator counts, component-level deltas.

\section{Results}

\paragraph{9.1 Detection efficacy and precision (RQ1)}
Headline validation metrics from \S4.5; Scenario~06 precision-stress confirmation (few/no findings on
a well-structured topology); Scenario~07 scalability confirmation.

\paragraph{9.2 Prescriptive efficacy (RQ2)}
SRI improves in all 7 scenarios (1.4\%--15.9\%), Wilcoxon $W=0.0$, $p=0.0156$.

\paragraph{9.3 Operator contributions (RQ3)}
Anti-affinity reallocation most frequent overall; QoS upgrades concentrate in high-loss profiles;
contribution counts cross-referenced against the anti-patterns each operator targets (\S6.3).

\paragraph{9.4 Remediation efficacy at component boundaries (RQ4)}
\textbf{Honest negative result}: mean $+4.61\%$ component-level improvement but regressions in
Hub-and-Spoke ($-31.67\%$) and Hyper-Scale ($-25.36\%$), traced to greedy unconditional policy
application; motivates the per-edit acceptance filter.

\paragraph{9.5 Computational overhead and CI/CD feasibility (RQ5)}
Detection gate sub-2s to $\sim$40s across scales; full generate--verify loop $<$65s for 6/7 scenarios,
$\sim$10.8 min at hyper-scale.

\section{Discussion and Threats to Validity}

\paragraph{10.1 What naming and verifying buys}
Over an unnamed criticality score --- \S9.1 and \S9.4 read together.

\paragraph{10.2 Positioning in CI/CD and technical-debt workflows}
Advisory prescriptions vs. blocking detection gate; the catalog as an architecture-review checklist
(design review by checklist, as in aviation/surgery) for teams without full CI/CD automation.

\paragraph{10.3 Construct validity}
Detection and verification both defined relative to the same discrete-event simulator; mitigation via
operators/patterns grounded in established dependability practice.

\paragraph{10.4 Internal validity}
Greedy, unconditional policy application; independence guarantee against circular leakage; stratified
reporting discipline (per-type and per-pattern, not only pooled).

\paragraph{10.5 External validity}
Synthetic parameterized topologies; catalog scope limited to the pub-sub communication paradigm
(hybrid REST/event architectures need additional patterns); QoS-mismatch detection currently specified
for DDS/ROS~2 and MQTT weight semantics.

\paragraph{10.6 Conclusion validity}
Small scenario sample ($n=7$ for the Wilcoxon test); single-run, verified-deterministic simulator
configuration.

\paragraph{10.7 Engineering trade-offs}
Of the verification step.

\section{Conclusion and Future Work}

\paragraph{11.1 Conclusions}
Summary of the end-to-end detect$\to$prescribe$\to$gate pipeline and headline results.

\paragraph{11.2 Future work, in priority order}
\begin{enumerate}[leftmargin=1.2cm]
  \item Per-edit acceptance filter --- highest priority, directly motivated by \S9.4.
  \item Budget-constrained policy search.
  \item Stochastic cascade propagation.
  \item Operator ablation and learned policy ordering.
  \item Catalog extension to hybrid REST/event and non-DDS/MQTT QoS semantics.
  \item Real-system replication (ATM topology).
  \item LLM-assisted PR generation --- kept to a single sentence, not oversold.
\end{enumerate}

\section*{References}

The merged reference list draws on the eleven previously verified prescriptive/SBSE/pub-sub citations
plus the anti-pattern catalog's own real, verified references (Fowler 1999; Brown et al. 1998;
Suryanarayana et al. 2014; Richardson 2018; Taibi et al. 2020; Baldwin \& Clark 2000; Lehman 1996;
Martin 2003; Nygard 2018; Colbourn 1987; Saaty 1980 --- deduplicated against the existing AHP
citation). One citation slot in \S2.3 (learning-based refactoring-opportunity mining) and one
(LLM-based refactoring suggestion) remain genuinely open and are \textbf{not} filled with invented
references.

\section*{Pre-Submission Checklist}

\begin{enumerate}
\item \textbf{Re-draft \texttt{draft.md} (the full manuscript) to match this outline} --- absorb
  \S4's catalog material into the manuscript body, renumber sections 5--11 accordingly, and reconcile
  the abstract/introduction/contributions list with the version in this file. This is now the single
  largest outstanding task; nothing below matters until the manuscript body exists in the new
  structure.
\item \textbf{Populate the two remaining \texttt{[REF: \ldots]} slots} in \S2.3 with real
  learning-based and LLM-based refactoring-recommendation citations. The smell-driven and
  architectural-technical-debt slots are now filled from \texttt{docs/antipatterns.md}'s own verified
  bibliography (Fowler 1999; Suryanarayana et al. 2014; Brown et al. 1998) --- no references have been
  invented for the remaining two.
\item \textbf{Literature pass:} expand from the current $\sim$20 references (11 prior + $\sim$9 newly
  merged from the catalog, after deduplicating Saaty) toward the $\sim$30--45 references AuSE
  reviewers typically expect.
\item \textbf{Confirm companion-paper status} ([1], the Software-as-a-Graph JSS submission, and the
  RASSE 2025 conference paper on graph-based critical-component identification) at submission time
  and disclose per AuSE's originality/overlap policy in the cover letter. Verify the RASSE 2025
  paper's scope does not overlap the newly added catalog material in \S4 beyond citation-level reuse
  of the shared RMAV/criticality foundation.
\item \textbf{Reconcile scenario counts explicitly in \S8.2} --- eight scenarios for detection
  validation vs. seven for prescriptive evaluation is intentional (per \texttt{docs/antipatterns.md}
  \S6.3's note) but must be stated plainly on first mention to avoid a reviewer inconsistency flag.
\item \textbf{Do not expand the LLM future-work mention} (\S11.2, item 7) beyond a single sentence ---
  the paper has no LLM/AutoML contribution, and overselling it risks a scope-mismatch desk rejection.
\item Fill in \textbf{Funding}, \textbf{Competing interests}, and \textbf{Data availability}
  declarations before submission.
\end{enumerate}

\end{document}
