\documentclass[review]{elsarticle}
% Compile with: pdflatex outline.tex (run twice for refs/ToC)
% "review" option = double-spaced, numbered lines, single column — standard for JSS submission.
% Swap to \documentclass[preprint,5p]{elsarticle} for camera-ready double-column layout.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}

\journal{Journal of Systems and Software}

\begin{document}

\begin{frontmatter}

\title{Graph Neural Networks for Reliability and Dependability Analysis in Complex Distributed Systems based on Publish--Subscribe Architecture}

\author{Author Name(s) Here}
\address{Affiliation Here}

\begin{abstract}
Publish-subscribe middleware decouples data producers and consumers, improving scalability but
obscuring the dependency chains along which a component's failure can cascade. Because runtime
telemetry does not exist before deployment, and static code analysis tools are blind to
system-level topology, identifying which components are critical, and why, remains difficult. We
address this gap with Software-as-a-Graph (SaG), a static system analysis framework that models a
publish-subscribe system as a typed, weighted, directed multigraph over five component classes and
derives logical dependencies through typed projection rules. On this representation, we train a
relation-specific heterogeneous graph neural network (GNN) with explicit quality-of-service (QoS)
edge-feature injection (HGL-QoS) to forecast each component's cascading failure impact before
deployment. We pair this learned predictor with an interpretable, Analytic Hierarchy
Process-weighted composite score decomposing criticality into four orthogonal quality dimensions,
so every diagnostic maps to a concrete remediation. Both predictors are validated against an
independent discrete-event cascade simulator under a strict input-label independence guarantee.
Across seven synthetic publish-subscribe topologies, the interpretable score suffices for known
architectures (rank correlation above 0.87), while the heterogeneous predictor is required to
generalize to unseen architectures (0.401 versus near zero for homogeneous baselines). A stratified
analysis confirms consistent predictive strength across component types, and an end-to-end
evaluation of prescriptive remediation operators yields a mixed result, exposing a gap between the
intended per-edit acceptance test and the current implementation. Finally, the framework operates
as a blocking continuous-integration and continuous-deployment (CI/CD) quality gate, evaluating
regressions in seconds with perfect precision and recall on injected structural faults.
\end{abstract}

\begin{highlights}
\item Heterogeneous GNN (HGL-QoS) generalizes cascade prediction to unseen architectures.
\item AHP-weighted composite score explains criticality via four orthogonal dimensions.
\item Typed multigraph model bridges the Architecture-Code Gap before deployment.
\item Delta-aware CI/CD gate blocks structural regressions in seconds, pre-deployment.
\end{highlights}

\begin{keyword}
heterogeneous graph neural networks \sep publish--subscribe systems \sep architectural
dependability \sep cascading failure \sep static system analysis \sep pre-deployment verification
\sep CI/CD quality gate
\end{keyword}

\end{frontmatter}

% ---------------------------------------------------------------------------
% MANUSCRIPT OUTLINE
% ---------------------------------------------------------------------------

\section{Introduction}

\begin{itemize}[leftmargin=*]
  \item \textbf{1.1 Motivation:} Rapid proliferation of pub-sub architectures (DDS, MQTT, ROS 2, Kafka) across high-availability cyber-physical, native cloud, and robotics domains. The core paradox: spatial, temporal, and synchronization decoupling that enables large-scale horizontal scaling simultaneously obscures logical failure-propagation paths. Hardening choices are cheapest and least disruptive pre-deployment—before any operational telemetry exists—exactly when systemic structural reasoning is most valuable.
  \item \textbf{1.2 The Architecture-Code Gap \& Problem Statement:} Formulates pre-deployment dependability analysis as two coupled sub-problems posed without execution traces: (i) \emph{quality attribution} (interpretable, multi-dimensional profiles mapping to distinct interventions) and (ii) \emph{failure-impact analysis} (quantifying cascade vulnerability). Defines the "Architecture-Code Gap" where clean, modular source code passing module-level checks can sit atop a fragile deployment topology riddled with single points of failure (SPOFs) or contract mismatches.
  \item \textbf{1.3 Limitations of Existing Approaches:} Investigates three central gaps: (1) Static Code Analysis (SCA) is blind to dynamic, system-level middleware topologies; (2) runtime chaos engineering carries operational risk and occurs too late to guide design; (3) topology-only and homogeneous learning-based centralities collapse typed semantics into flattened views, leading to representation collapse.
  \item \textbf{1.4 Our Approach \& Research Questions:} Introduces Software-as-a-Graph's (SaG) typed multigraph, multi-dimensional quality attribution, dual predictors, prescriptive counterfactual remediation, and automated CI/CD gating using a thread-safe, database-free in-memory repository to eliminate live-database build dependencies. Details the four primary research questions:
    \begin{description}[leftmargin=2.2cm,style=nextline]
      \item[RQ1 (Interpretable vs. Learned)] When does deterministic multi-attribute attribution suffice for criticality prediction, and when is a trained relation-specific heterogeneous graph neural network required to recover simulated cascade orderings on unseen architectures?
      \item[RQ2 (Typing \& Structural Anomalies)] What failure modes does multi-dimensional, typed structural attribution expose that a single-score topological centrality conceals?
      \item[RQ3 (QoS Impact)] How does explicit multi-attribute QoS contract edge-feature injection affect in-distribution convergence versus out-of-distribution Leave-One-Scenario-Out (LOSO) generalizability?
      \item[RQ4 (CI/CD Operationalization)] What is the practical feasibility, scaling behavior, and performance overhead of executing this analysis as a blocking build gate in continuous integration pipelines?
    \end{description}
  \item \textbf{1.5 Contributions:} (1) Typed system graph model integrating code-level SCA metrics; (2) automated, delta-aware CI/CD Quality Gate; (3) pre-deployment failure-impact analysis under a strict view-based independence guarantee ($Q(v)$ and $HGL\text{-}QoS$); (4) prescriptive remediation framework providing counterfactual evaluation with an honestly disclosed implementation gap; (5) stratified, type-specific tracking of predicted-to-simulated correlation.
  \item \textbf{1.6 Relationship to the Authors' Prior Work:} Consolidates an initial structural baseline layer projection framing [Anon-A] with the relation-specific graph transformer predictor, multi-dimensional attribution, and continuous gating. Explicitly states that no companion predictor paper is under duplicate review, and clarifies that placeholder domain-expert panel case studies from early configurations have been fully withdrawn.
  \item \textbf{1.7 Organization:} Roadmap detailing subsequent manuscript sections.
\end{itemize}

\section{Related Work}

\begin{itemize}[leftmargin=*]
  \item \textbf{2.1 Publish--Subscribe Middleware and Dependability:} Outlines traditional protocol-level fault tolerance, reliable dissemination overlays, and replication schemes in DDS/MQTT/Kafka systems, positioning SaG as an earlier lifecycle intervention (design-time verification vs.\ runtime mitigation).
  \item \textbf{2.2 Static Code Analysis (SCA) vs. Static System Analysis (SSA):} Tracks expansion from intra-component AST parsing toward global, inter-component SSA topologies that ingest SCA metrics as node attributes rather than replacing them.
  \item \textbf{2.3 Continuous Pre-Deployment Verification and Gating:} Evaluates chaos engineering limits versus shifting verification left via Architecture-as-Code (AaC). Links system delta-gating to the "Clean as You Code" paradigm utilized by mature code gates.
  \item \textbf{2.4 Structural Criticality Analysis:} Discusses classical centralities, articulation points, and cascading network disruptions. Notes their core limitation: dimensional collapse and blindness to the shared-library simultaneous blast mode.
  \item \textbf{2.5 Learning-Based Criticality Prediction:} Evaluates graph learning centralities (FINDER, DrBC, PowerGraph) and details why homogeneous architectures collapse typed semantics, making the case for relation-specific message passing (RGCN, HAN, HGT, MAGNN) to prevent over-smoothing in hub-heavy networks.
  \item \textbf{2.6 Quality Attributes and Multi-Criteria Scoring:} Explores software quality attribution via decision theory. Contextualizes the unique use of AHP as an explicit, auditable attribution and explanation system rather than a generic prioritization tool.
  \item \textbf{2.7 Architectural Remediation and Anti-Pattern Detection:} Contrasts static refactoring heuristics with SaG's closed-loop, simulation-backed counterfactual testing framework.
  \item \textbf{2.8 Positioning:} Synthesizes the exact five-way gap (runtime-only, code-only, untyped-structural, untyped-learned, prioritization-only) closed jointly by the framework.
\end{itemize}

\section{The Software-as-a-Graph (SaG) Model}

\begin{itemize}[leftmargin=*]
  \item \textbf{3.1 Nodes, Edges, and the Formal Object:} Defines the multigraph object $G = (V, E, \tau_V, \tau_E, w_E, w_V)$. Partitions $V$ into five distinct node types (Applications, Brokers, Topics, Nodes, Libraries) and maps $E$ onto six directional structural relations (\texttt{PUBLISHES\_TO}, \texttt{SUBSCRIBES\_TO}, \texttt{ROUTES}, \texttt{RUNS\_ON}, \texttt{CONNECTS\_TO}, \texttt{USES}).
  \item \textbf{3.2 QoS-Aware Edge and Vertex Weights:} Details the two-stage allocation of contract coupling strength:
    $$\text{QoS\_score} = 0.30r + 0.40d + 0.30p$$
    $$\text{size\_norm} = \min\!\left(\frac{\log_2(1 + \text{size\_kb})}{50},\ 1.0\right)$$
    $$w(e) = \beta\cdot\text{QoS\_score} + (1-\beta)\cdot\text{size\_norm}, \qquad \beta = 0.85, \text{ floor } 0.01$$
    Maps symbolic reliability ($r$), durability ($d$), and priority ($p$) parameters via AHP weights. Outlines vertex weight ($w_V$) aggregation rules, including the log-scale fan-out-amplified library weighting.
  \item \textbf{3.3 Derived Dependencies --- the \texttt{DEPENDS\_ON} Projection:} Formalizes six type-specific projection rules (\texttt{app\_to\_app}, \texttt{app\_to\_broker}, etc.) directed from dependent to dependency. Formulates multi-topic edge consolidation tracking $\max w(t)$ and path multiplicity via \texttt{path\_count}. Contrasts the step-by-step propagation of Rule 1 (\emph{sequential cascade}) with the single-event impact of Rule 5 (\emph{simultaneous blast}).
  \item \textbf{3.4 Ingestion of Code-Level SCA Metrics:} Details the ingestion of SonarQube \texttt{cm\_*} vertex attributes (LOC, WMC, LCOM, CBO, RFC, technical debt ratio, bugs, vulnerabilities) into Application and Library components to build the Code Quality Penalty.
  \item \textbf{3.5 Graph Views and Multi-Layer Projections:} Introduces the load-bearing view separation: $G_{\text{structural}}$ (simulator oracle) vs.\ $G_{\text{analysis}}(\ell)$ (analytics engine), establishing the non-circular independence guarantee. Outlines the four analytical projection layers (Application, Infrastructure, Middleware, System) and their mapping to a hierarchical MIL-STD-498 reporting rollup.
  \item \textbf{3.6 Running Example:} Contextualizes a three-application, one-topic, one-shared-library scenario illustrating graph generation, logical projections, and failure behavior.
\end{itemize}

\section{Multi-Dimensional Quality Attribution (The Interpretable Path)}

\begin{itemize}[leftmargin=*]
  \item \textbf{4.1 Four Orthogonal Dimensions:} Formulates the Reliability, Maintainability, Availability, and Vulnerability (RMAV) framework. Proves that each structural graph metric is allocated to exactly one dimension by design, ensuring structural profiles directly reflect actionable architectural deficiencies.
  \item \textbf{4.2 RMAV Formulas:} Complete formal mathematical presentation of the dimension metrics:
    \begin{align*}
      R(v) &= 0.45\,\mathrm{RPR}(v) + 0.30\,\mathrm{DG\_in}(v) + 0.25\,\mathrm{CDPot\_enh}(v) \quad [\tau_V(v)\neq\text{Topic}] \\
      M(v) &= 0.35\,\mathrm{BT}(v) + 0.30\,w\_\text{out}(v) + 0.15\,\mathrm{CQP}(v) + 0.12\,\mathrm{CouplingRisk\_enh}(v) + 0.08\,(1-\mathrm{CC}(v)) \\
      A(v) &= 0.35\,\mathrm{AP\_c\_directed}(v) + 0.25\,\mathrm{QSPOF}(v) + 0.25\,\mathrm{BR}(v) + 0.10\,\mathrm{CDI}(v) + 0.05\,w(v) \\
      V(v) &= 0.40\,\mathrm{REV}(v) + 0.35\,\mathrm{RCL}(v) + 0.25\,w\_\text{in}(v)
    \end{align*}
    Defines Reverse PageRank ($\mathrm{RPR}$), Enhanced Cascade Potential ($\mathrm{CDPot\_enh}$), Betweenness Centrality ($\mathrm{BT}$), Code Quality Penalty ($\mathrm{CQP}$), Directed Articulation score ($\mathrm{AP\_c\_directed}$), Bridge Ratio ($\mathrm{BR}$), and the Connectivity Degradation Index ($\mathrm{CDI}$).
  \item \textbf{4.3 The Composite Score $Q(v)$:} Blends raw matrix-derived AHP weights $(0.43, 0.24, 0.17, 0.16)$ satisfying $\mathrm{CR}<0.02$ with a uniform prior via shrinkage parameter $\lambda=0.70$, yielding operational scoring weights: $Q(v) = 0.38\,A(v) + 0.24\,R(v) + 0.19\,M(v) + 0.19\,V(v)$.
  \item \textbf{4.4 Adaptive Criticality Classification:} Establishes box-plot fence categorization ($Q_3 + 1.5\,\mathrm{IQR}$) to dynamically map scores into \texttt{CRITICAL}/\texttt{HIGH}/\texttt{MEDIUM}/\texttt{LOW}/\texttt{MINIMAL} severity flags, utilizing a percentile fallback system for sparse graph configurations ($n < 12$).
  \item \textbf{4.5 Determinism and the Independence Guarantee:} Formal confirmation that $Q(v)$ inputs derive strictly from deterministic structural passes over $G_{\text{analysis}}$, entirely isolated from the stochastically seeded simulation oracle.
  \item \textbf{4.6 Worked Attribution:} Maps §3.6's running example onto the equations to demonstrate clear, non-overlapping profile shapes for an Availability-centric SPOF vs.\ a Reliability-centric cascade hub.
\end{itemize}

\section{Failure-Impact Analysis via Heterogeneous GNN and Interpretable Forecasting}

\begin{itemize}[leftmargin=*]
  \item \textbf{5.1 Ground-Truth Impact $I(v)$:} Outlines the discrete-event failure simulator operating on $G_{\text{structural}}$ to measure cascade loss as an AHP-weighted compound: $I(v) = 0.35\,\text{reachability\_loss} + 0.25\,\text{fragmentation} + 0.25\,\text{throughput\_loss} + 0.15\,\text{flow\_disruption}$. Details parameters: $0.2$ propagation threshold, 10-epoch horizon, and 5-seed averaging.
  \item \textbf{5.2 Two Predictors over the Same Model:} Compares the deterministic $Q(v)$ model with inductive Graph Transformers parameterized by node/edge type, contrasting the QoS-masked base (**HGL**) model with the QoS-encoded architecture (**$HGL\text{-}QoS$**).
  \item \textbf{5.3 The Independence Guarantee:} Explains view-based property boundaries preventing simulation performance or label states from entering GNN node layers.
  \item \textbf{5.4 The Shared-Library Blast Mechanism --- A Negative Result:} Documents an explicit empirical evaluation across 165 shared library components that disproved the low-$Q$/high-$I$ under-reporting hypothesis. Proves $I(v) \le Q(v)$ globally ($Q_{\max} = 0.422, I = 0.086$) and notes single-component impact caps at $0.320$, presenting this as an honest negative result.
  \item \textbf{5.5 Stratified Correlation --- A Consistency Check:} Reports pooled Spearman correlation ($\rho = 0.374$, $p \approx 2.2\times10^{-52}$) across 1,545 nodes. Evaluates stratified inside-class bounds ($\rho = 0.322$--$0.429$), ruling out Simpson's-paradox masking and proving consistent predictive content across all component types.
\end{itemize}

\section{Prescriptive Remediation and CI/CD Quality Gating}

\begin{itemize}[leftmargin=*]
  \item \textbf{6.1 A Two-Phase Generate--Verify Procedure:} Splits candidate selection from performance estimation: Generate operates on structure/attribution, while Verify re-runs the simulation oracle on mutated counterfactual subgraphs $G' = e(G)$.
  \item \textbf{6.2 Remediation Operators:} Formalizes four operators: \texttt{RedundancyInsertion}, \texttt{PathDiversification}, \texttt{FanOutReduction}, and \texttt{SharedTopicReduction}, each paired with specific structural triggers and target quality dimensions.
  \item \textbf{6.3 Triggering on Blast Radius, not $Q(v)$:} Proves why \texttt{FanOutReduction} fires directly on structural fan-out and consumer multiplicity rather than $Q(v)$ to bypass library score dilution.
  \item \textbf{6.4 Acceptance Criterion (Target Design) and the Open Implementation Gap:} Establishes the intended robust acceptance rule ($\Delta I > \kappa\,\sigma_{\text{seed}}$ across the threshold sweep). Discloses the current implementation gap where \texttt{PrescribeService} applies mutations unconditionally without a per-edit counterfactual filter.
  \item \textbf{6.5 Independence Invariants:} Enforces the three operational invariants ensuring candidate selections do not feed back or cause circular optimizations during execution.
  \item \textbf{6.6 CI/CD Quality Gate Implementation:} Details \texttt{detect\_antipatterns.py} build integration using delta semantics against the merge base, a waiver register for accepted SPOFs, the three-tier exit-code protocol (0=pass, 1=warning, 2=blocked), and the role of the database-free \texttt{MemoryRepository} in eliminating live database transaction overhead.
  \item \textbf{6.7 What Remediation Currently Yields --- A Mixed Result:} Presents an end-to-end impact table showing a positive mean cross-scenario improvement of $+4.61\%$ that conceals severe resilience regressions in 3 of 7 systems (Hub-and-Spoke $-31.67\%$, Financial Trading $-27.66\%$, Enterprise $-25.36\%$), demonstrating the direct consequence of the §6.4 implementation gap.
\end{itemize}

\section{Experimental Setup}

\begin{itemize}[leftmargin=*]
  \item \textbf{7.1 Datasets:} Details the seven synthetic scenarios comprising 1,545 nodes (850 Apps, 375 Topics, 165 Libraries, 119 Nodes, 36 Brokers) across scale presets up to \texttt{xlarge}, noting that the Early expert-ranking real-world datasets have been entirely purged.
  \item \textbf{7.2 Predictors and Baselines:} Documents the verification matrix: RMAV/$Q$, HGL, $HGL\text{-}QoS$, homogeneous GL/GL-QoS baselines, and structural centrality baselines Topo-BL/Topo-QoS.
  \item \textbf{7.3 Evaluation Metrics:} Primary metrics include Spearman $\rho$, NDCG@10, Top-K overlap, F1, and SPOF-F1 scores, backed by bootstrap intervals ($B=2000$) and paired Wilcoxon tests under validation targets of $\rho \ge 0.70$ and $F_1 \ge 0.80$.
  \item \textbf{7.4 Protocols:} Compares in-distribution per-scenario splits against inductive Leave-One-Scenario-Out (LOSO) cross-validation; tracks performance across five fixed seeds.
  \item \textbf{7.5 Canonical Simulator and Reproducibility:} Explicitly sets environment constants: step blast semantics, propagation threshold $= 0.2$, 10-epoch horizon, and seeds $\{42, 123, 456, 789, 2024\}$.
\end{itemize}

\section{Results}

\begin{itemize}[leftmargin=*]
  \item \textbf{8.1 RQ1 --- Interpretable Attribution vs. Learning:} Reports that in-distribution, $Q(v)$ attribution is strongly aligned ($\rho > 0.87$, $F_1 > 0.90$), outperforming learned alternatives. Out-of-distribution (LOSO), $HGL\text{-}QoS$ decisively dominates ($\rho = 0.401$ vs.\ $\approx 0$ for homogeneous variants), proving learning is required for cross-architecture transfer. Notes measurement context caveats.
  \item \textbf{8.2 RQ2 --- What Taking Type Seriously Shows (and Does Not Show):} Quantifies typing-based performance gains ($\Delta F_1 = +0.284$ in-distribution, $\Delta\rho = +0.286$ LOSO), reviews the library blast null result, and re-confirms stratified profile consistency.
  \item \textbf{8.3 RQ3 and Robustness --- Ablations and Sensitivity:} Reports an in-distribution null result for QoS encoding vs.\ its major out-of-distribution transfer benefit ($\rho = 0.401$ vs.\ $0.307$). Evaluates AHP shrinkage ($\lambda$) and propagation-threshold parameter sweeps.
  \item \textbf{8.4 RQ4 --- Feasibility and Performance of the CI/CD Quality Gate:} Documents execution footprints proving sub-quadratic scaling via the \texttt{MemoryRepository} ($<2\,\text{s}$ tiny, $\approx 5\,\text{s}$ medium, $\approx 40\,\text{s}$ xlarge xlarge). Confirms a perfect 1.0/1.0 precision/recall on injected regression blocks.
\end{itemize}

\section{Discussion, Threats to Validity, and Conclusion}

\begin{itemize}[leftmargin=*]
  \item \textbf{9.1 Interpretation:} Synthesizes the clear complementary division of labor between attribution and learning, the value of structural typing, open-loop implementation realities, and shift-left continuous gating pipelines.
  \item \textbf{9.2 Threats to Validity:} Explicitly catalogs threats: Construct (simulation limits), Internal (independence guarantees against circular testing), and External (synthetic boundaries).
  \item \textbf{9.3 Limitations and Future Work:} Outlines 5 concrete next steps: (1) unified per-scenario head-to-head tracking split matrix; (2) trust boundary security proxies; (3) real-world/expert-panel practitioner validation studies; (4) building the missing per-edit counterfactual filter inside \texttt{PrescribeService}; (5) production incident trace calibration.
  \item \textbf{9.4 Conclusion:} Closing synthesis: Software-as-a-Graph successfully bridges the Architecture-Code Gap pre-deployment, reporting negative and mixed findings honestly to ensure legible, actionable design-time assurance.
\end{itemize}

\section*{References}

\begin{itemize}[leftmargin=*]
  \item 15 numbered references spanning pub-sub foundations [1--3], network-science centrality and
    cascading-failure theory [4--6], learning-based key-node identification and benchmarks [7--9],
    heterogeneous GNN architectures [10--13], over-smoothing [14], and AHP [15]. All entries
    verified as real, published works, explicitly confirming the NeurIPS 2024 Track history of
    \emph{PowerGraph} [9].
\end{itemize}

\end{document}